---
title: "antclassify_workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{antclassify_workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(AntClassify)
```

## Introduction

The **AntClassify** package provides an integrated ecological pipeline to classify ant communities into functional guilds, identify exotic species, detect endemic and rare species of the Atlantic Forest, and quantify key ecological patterns.

This tool was developed to facilitate ecological analyses, standardize functional classification, and improve reproducibility in ant community studies. By integrating multiple ecological databases into a single workflow, AntClassify allows researchers to efficiently assess community structure, biological invasions, endemism, and rarity patterns.

The package is particularly useful for biodiversity monitoring, conservation planning, and macroecological research involving ant assemblages.

AntClassify aims to provide a standardized and reproducible framework for advancing ecological research on ant communities.

## Example dataset

To demonstrate the functionality of the package, we create a more realistic ant community dataset.

```{r}
dados <- data.frame(
  Atta_sexdens = 50,
  Camponotus_atriceps = 40,
  Crematogaster_sp = 35,
  Cyphomyrmex_minutus = 30,
  Cyphomyrmex_rimosus = 28,
  Ectatomma_edentatum = 25,
  Heteroponera_mayri = 22,
  Holcoponera_striatula = 20,
  Monomorium_floricola = 18,
  Monomorium_pharaonis = 17,
  Pheidole_megacephala = 16,
  Strumigenys_emmae = 15,
  Strumigenys_rogeri = 14,
  Nylanderia_fulva = 13,
  Odontomachus_chelifer = 12,
  Oxyepoecus_reticulatus = 11,
  Pachycondyla_striata = 10,
  Apterostigma_serratum = 9,
  Brachymyrmex_delabiei = 8,
  Brachymyrmex_feitosai = 7,
  Camponotus_fallatus = 6,
  Camponotus_hermanni = 5,
  Camponotus_xanthogaster = 4,
  Pheidole_aberrans = 3,
  Pheidole_fimbriata = 3,
  Pheidole_obscurithorax = 2,
  Pheidole_subarmata = 2,
  Strumigenys_fridericimuelleri = 2,
  Heteroponera_inermis = 2,
  Oxyepoecus_browni = 2,
  Sphinctomyrmex_stali = 1,
  Strumigenys_sanctipauli = 1,
  Brachymyrmex_micromegas = 1,
  Camponotus_tripartitus = 1,
  Diaphoromyrma_sofiae = 1
)

colnames(dados) <- gsub("_", " ", colnames(dados))

dados
```

## Running the pipeline

The main function of the package, `antclassify()`, runs the full ecological workflow.

```{r}
resultado <- antclassify(dados)
```

## Accessing results

You can access each component of the results:

```{r}
names(resultado)
head(resultado$guilds$table)
```
